%\documentclass[12pt,a4paper]{amsart}
\documentclass[11pt,a4paper]{article}
%\ifx\pdfpageheight\undefined\PassOptionsToPackage{dvips}{graphicx}\else%
%\PassOptionsToPackage{pdftex}{graphicx}
\PassOptionsToPackage{pdftex}{color}
%\fi

%\usepackage{diagrams}

%\usepackage[all]{xy}
\usepackage{url}
\usepackage[utf8]{inputenc}
\usepackage{verbatim}
\usepackage{latexsym}
\usepackage{amssymb,amstext,amsmath,amsthm}
\usepackage{epsf}
\usepackage{epsfig}
% \usepackage{isolatin1}
\usepackage{a4wide}
\usepackage{verbatim}
\usepackage{proof}
\usepackage{latexsym}
%\usepackage{mytheorems}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{corollary}{Corollary}[theorem]
\newtheorem{lemma}{Lemma}[theorem]
\newtheorem{proposition}{Proposition}[theorem]
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{remark}{Remark}[theorem]
\newtheorem{TODO}{TODO}[theorem]

\usepackage{float}
\floatstyle{boxed}
\restylefloat{figure}


%%%%%%%%%copied from SymmetryBook by Marc

% hyperref should be the package loaded last
%% \usepackage[backref=page,
%%             colorlinks,
%%             citecolor=linkcolor,
%%             linkcolor=linkcolor,
%%             urlcolor=linkcolor,
%%             unicode,
%%             pdfauthor={CAS},
%%             pdftitle={Symmetry},
%%             pdfsubject={Mathematics},
%%             pdfkeywords={type theory, group theory, univalence axiom}]{hyperref}
% - except for cleveref!
\usepackage[capitalize]{cleveref}
%\usepackage{xifthen}
\usepackage{xcolor}
\definecolor{linkcolor}{rgb}{0,0,0.5}

%%%%%%%%%
\def\oge{\leavevmode\raise
.3ex\hbox{$\scriptscriptstyle\langle\!\langle\,$}}
\def\feg{\leavevmode\raise
.3ex\hbox{$\scriptscriptstyle\,\rangle\!\rangle$}}

%%%%%%%%%
\newcommand\myfrac[2]{
 \begin{array}{c}
 #1 \\
 \hline \hline
 #2
\end{array}}


\newcommand*{\Scale}[2][4]{\scalebox{#1}{$#2$}}%
\newcommand*{\Resize}[2]{\resizebox{#1}{!}{$#2$}}

\newcommand{\II}{\mathbb{I}}
\newcommand{\refl}{\mathsf{refl}}
\newcommand{\mkbox}[1]{\ensuremath{#1}}


\newcommand{\Id}{\mathsf{Id}}
\newcommand{\conv}{=}
%\newcommand{\conv}{\mathsf{conv}}
\newcommand{\lam}[2]{{\langle}#1{\rangle}#2}
\def\NN{\mathsf{N}}
\def\UU{\mathsf{U}}
\def\JJ{\mathsf{J}}
\def\Level{\mathsf{Level}}
\def\List{\mathsf{List}}
\def\Cons{\mathsf{Cons}}
\def\Nil{\mathsf{Nil}}
%\def\Type{\hbox{\sf Type}}
\def\ZERO{\mathsf{0}}
\def\SUCC{\mathsf{S}}

\newcommand{\type}{\mathsf{type}}
\newcommand{\LAM}{\lambda}
\newcommand{\APP}{\mathsf{app}}
\newcommand{\mypi}[3]{\Pi_{#1:#2}#3}
\newcommand{\mylam}[3]{\lambda_{#1:#2}#3}
\newcommand{\mysig}[3]{\Sigma_{#1:#2}#3}
\newcommand{\N}{\mathsf{N}}
\newcommand{\Set}{\mathsf{Set}}
\newcommand{\El}{\mathsf{El}}
%\newcommand{\U}{\mathsf{U}} clashes with def's in new packages
\newcommand{\T}{\mathsf{T}}
\newcommand{\sT}{\mathsf{t}}
\newcommand{\Usuper}{\UU_{\mathrm{super}}}
\newcommand{\Tsuper}{\T_{\mathrm{super}}}
%\newcommand{\conv}{\mathrm{conv}}
\newcommand{\idtoeq}{\mathsf{idtoeq}}
\newcommand{\isEquiv}{\mathsf{isEquiv}}
\newcommand{\ua}{\mathsf{ua}}
\newcommand{\UA}{\mathsf{UA}}
%\newcommand{\Level}{\mathrm{Level}}
\def\Constraint{\mathsf{Constraint}}
\def\Ordo{\mathcal{O}}

\newcommand{\Con}{\mathsf{ Con}}
\newcommand{\Elem}{\mathsf{Elem}}
\newcommand{\Type}{\mathsf{Type}}
\newcommand{\id}{\mathsf{id}}
\newcommand{\pp}{\mathsf{p}}
\newcommand{\qq}{\mathsf{q}}

\def\Ctx{\mathrm{Ctx}}
\def\Ty{\mathrm{Ty}}
\def\Tm{\mathrm{Tm}}

\def\CComega{\mathrm{CC}^\omega}
\setlength{\oddsidemargin}{0in} % so, left margin is 1in
\setlength{\textwidth}{6.27in} % so, right margin is 1in
\setlength{\topmargin}{0in} % so, top margin is 1in
\setlength{\headheight}{0in}
\setlength{\headsep}{0in}
\setlength{\textheight}{9.19in} % so, foot margin is 1.5in
\setlength{\footskip}{.8in}

% Definition of \placetitle
% Want to do an alternative which takes arguments
% for the names, authors etc.

\newcommand{\natrec}{\mathsf{natrec}}
%\rightfooter{}
\newcommand{\set}[1]{\{#1\}}
\newcommand{\sct}[1]{[\![#1]\!]}
%\usepackage{diagrams}
\usepackage{color}
\newcommand\coloremph[2][red]{\textcolor{#1}{\emph{#2}}}
\newcommand\norm[1]{\left\lVert #1 \right\rVert}
\newcommand\greenemph[2][green]{\textcolor{#1}{\emph{#2}}}
\newcommand{\EMP}[1]{\emph{\textcolor{red}{#1}}}




\begin{document}

\title{Type Theory with a Cumulative Hierarchy of Universes}

\author{}
\date{}
\maketitle

%\begin{abstract}
%\end{abstract}

\section{Introduction}\label{sec:intros}

 There are two different ways to formulate the notion of universe in type theory. Martin-Löf's first versions \cite{martinlof:72,martinlof:predicative,martinlof:hannover,martinlof:padova} were formulated \`a la Russell, where a universe is seen as a type of types. Later \cite{martinlof:padova}, an alternative version \`a la Tarski, where a universe is seen as a type of names of types and comes with a decoding function, was also introduced.
 
Some recent works \cite{Assaf14,Thire20} show the equivalence between these two formulations, but these arguments do not apply to formulation of type theory with typed equality judgements \cite{martinlof:hannover,martinlof:padova}. In this paper we show the equivalence for this case, using a method due to Streicher  \cite{Streicher:semtt} that seems simpler
and more transparent than the one in \cite{Assaf14,Thire20}.

We quote from Streicher, p 246, concerning his theorem 4.10.
\begin{quotation}
This theorem, although interesting in itself, has in particular the consequence that type information in application terms is redundant, i.~e.~if $t$ and $s$ are terms which are different only w.r.t. the type information of subterms which are application terms then $t$ and $s$ are provably equal.
\end{quotation}
The formal statement is theorem 4.11

Streicher's argument relies
on  basic properties of such a type system (such as subject reduction and normalization). Such properties were assumed
in \cite{Streicher:semtt}, but they are now known to be provable
for systems with a predicative hierarchy of universes \cite{coquand:sophia,abel:wroclaw,coq18}.

The argument is as follows. There is a natural forgetful map from the Tarski version to the Russell version. Modulo the (known) meta-properties of both system, one can show that any Russell judgement can be lifted in a unique way to a Tarski judgement.\footnote{PD: Cf the sketch of a proof using the initial $\Pi\UU$-cwf of normal forms in {\tt initial-cwf-normal-forms.tex} in the same directory.}

\section{An annotated type theory with Tarski style universes}

We now present the system $T_1$ of type theory with $\Pi$-types with annotated application and abstraction and an external hierarchy of Tarski style universes $\UU_n$ with decoding maps $\T_n$. For simplicity we only consider closure under $\Pi$-types. 

The syntax\footnote{PD: I suggest being consistent about whether to write e g $\T_n a$ or $\T_n(a)$, etc, and vote for the latter} for types is
$$
A,B~::=~\Pi~{A}~B~|~\UU_n~|~\T_n a
$$
and the syntax for terms is
$$
a,b~::=~v_i~|~\APP(A,B,c,a)~|~\LAM(A,B,b)~|~\Pi^n~ a~b~|~\UU^k_n~|~\sT^k_n a~
$$
Note that we present a system with de Bruijn variables and implicit substitution.
Here $v_i$ is a variable with de Bruijn index $i$. In this system, applications $\APP(A,B,c,a)$ and abstractions $\LAM(A,B,b)$ are annotated with the argument type $A$ and the result type family $B$. Moreover, $\Pi^n~ a~b$ are codes for $\Pi$-types in $\UU_n$, $\UU^k_n$ is the code for $\UU_k$ in $\UU_n$, where $k < n$, and $\sT^k_n a$ is the lifting of an element $a$ in $\UU_k$ to $\UU_n$. We thus consider a cumulative hierarchy of universes.

The theory $T_1$ is set up to facilitate the construction of an initial category with families with extra structure for $\Pi$-types and a cumulative hierarchy of universes, see appendix. (We also need to define equality of contexts $\Gamma = \Gamma'$ and $\gamma = \gamma' : \Delta \to \Gamma$.) A proof of a similar construction has been formalized in Agda by Brunerie et al \cite{brunerie:initial}, who proved that a similar calculus is initial in a category of contextual categories with suitable extra structure. Note that initial categories with families (and contextual categories) with suitable extra structure can also be constructed in an alternative way from a calculus of explicit substitutions  \cite{casellan:tlca15}.

In the following section we will define the theory $T_2$ of type theory with $\Pi$-types and a cumulative hierarchy of universes \`a la Russell. In this system the application $c\,a$ is not annotated and the abstraction n$\lambda A b$ is only annotated with the argument type $A$ and not the result family $B$. We will prove that $T_1$ and $T_2$ are isomorphic 
category with families with extra structure for $\Pi$-types and a cumulative hierarchy of universes, and thus the $T_2$ is initial in this category too.
\medskip

We define weakenings\footnote{PD: changed from "renamings"} $r$ and one place substitutions $\sigma$:
$$
r~::=~\pp~|~r^+~~~~~~~~~\sigma~::=~[a]~|~\sigma^+~~~~~~~~~~\alpha~::=~r~|~\sigma
$$
We first define the action of a weakening on types $Ar$ and terms $ar$ and then the action of a substitution on types $A\sigma$ and terms $a\sigma$ by induction on $A$ and $a$. 

The actions on variables are
$$
v_i\pp = v_{i+1}~~~~~v_0\alpha^+ = v_0~~~v_{i+1}\alpha^+ = (v_i\alpha)\pp~~~~~v_0[a] = a~~~v_{i+1}[a] = v_i\footnote{PD: added}
$$
We can then define $A\alpha$ (resp. $a\alpha$) by induction on $A$ (resp. $a$).
For instance, $(\Pi~A~B)\alpha = \Pi~(A\alpha)~(B\alpha^+)$
and $\APP(A,B,c,a)\alpha = \APP(A\alpha,B\alpha^+,c\alpha,a\alpha)$.

\begin{lemma}
  We have $(B[a])\alpha = B\alpha^+[a\alpha]$ and $(b[a])\alpha = b\alpha^+[a\alpha]$.
\end{lemma}

\medskip

The type system describes how to derive judgements of the form $\Gamma\vdash$ and $\Gamma\vdash J$ where $J$ is of the form
$A~\type$ or $a:A$ or $A=B$ or $a=b : A$.

\medskip

The judgment $\Gamma\vdash$ expresses that $\Gamma$ is a well-formed context. $\Gamma\vdash A$ expresses that $A$ is a type in context $\Gamma$.
We may write it $A~\type~(\Gamma)$ and may omit the global context $\Gamma$.

First, we have the following general typing rules. These are common to the present system and the Russell-style system to be presented later:
\begin{figure}
  \caption{Context formation rules and typing rules for variables}
$$
\frac{\Gamma\vdash A}{\Gamma.A\vdash}~~~~~~\frac{}{()\vdash}~~~~~~~
\frac{\Gamma\vdash A}{\Gamma.A\vdash v_0:A\pp}~~~~~~~
\frac{\Gamma\vdash v_i:A~~~~~~~\Gamma\vdash B}{\Gamma.B\vdash v_{i+1}:A\pp}~~~~~~~
$$
\end{figure}
The typing rules for $\Pi$-types and universes are shown in Figure \ref{type-Tarski}.
\footnote{PD: I have restructured the rules a little. First, we can show the general rules which are common to both systems. Then the rules for annotated $\Pi$-types. And finally, the rules for Tarski-style universes. Then we only need to show the rules for non-annotated $\Pi$-types and the rules for Russell-style universes in the following section.}

\begin{figure}
  \caption{Typing rules for annotated $\Pi$-types and Tarski-style universes}\label{type-Tarski}
%$$
%  \frac{A~\type}{v_0:A\pp~(A)}
%  ~~~~~~
%  \frac{v_i:A~~~~B~\type}{v_{i+1}:A\pp~(B)}
%$$
$$
    \frac{A~\type~~~~~~B~\type~(A)}{\Pi~A~B~\type}
 ~~~~~~~~~
\frac{b:B~(A)}{\LAM(A,B,b):\Pi~A~B}
~~~~~~~~
\frac{c:\Pi~A~B~~~~~~a:A}
     {\APP(A,B,c,a):B[a]}
$$
$$
\frac{}{\UU_n~\type}~~~~~~
\frac{a:\UU_{n}}{\T_{n}(a)~\type}~~~~~~
\frac{a:\UU_{n}~~~~~~b:\UU_n~(\T_{n}(a))}
     {\Pi^{n} a b:\UU_{n}}
        ~~~~~~~
\frac{}{\UU^{n}_m:\UU_{n}}m<n
~~~~~~~~~\frac{a:\UU_{m}}{\sT_{m}^{n}(a):\UU_{n}}m<n
$$
\end{figure}

We write $\conv$ for definitional equality (or conversion).
The main conversion rules are shown in Figure \ref{conv-Tarski}.

\begin{figure}
  \caption{Conversion rules for Tarski-style}\label{conv-Tarski}
$$
\frac{ a:A~~~~~~ A~ \conv~ B}{ a:B}~~~~~~~~~
\frac{ a ~\conv~a':A~~~~~~ A  ~\conv~ B}{ a ~\conv~a':B}
$$
$$
\frac{A~=~B~~~~~A~=~C}{B~=~C}~~~~~~~~~\frac{A~\type}{A~=~A}~~~~~~~~~
\frac{a~=~b:A~~~~~a~=~c:A}{b~=~c:A}~~~~~~~~~\frac{a:A}{a~=~a:A}
$$
$$
\frac{A~=~A'~~~~~~B~=~B'~(A)}{\Pi~A~B~=~\Pi~A'~B'}~~~~~~~~
\frac{c~=~c':\Pi~A~B~~~~~~a~=~a':A~~~~~~A~=~A'~~~~~~~~B~=~B'}{\APP(A,B,c,a)~=~\APP(A',B',c',a'):B[a]}
$$
%$$
%\frac{A~=~A'~~~~~~B~=~B'~(A)~~~~~b~=~b':B~(A)}{\LAM(A,B,b)~=~\LAM(A',B',b'):\Pi~A~B}
%$$
$$
\frac{b:B~(A)~~~~~~~~ a:A}{ \APP(A,B,\LAM(A,B,b),a)  ~\conv~ b[a]:B[a]}
~~~~~~~
\frac{\APP(A,B,f\pp,v_0) = \APP(A,B,g\pp,v_0):B~(A)}{ f = g : \Pi~A~B}\footnote{PD: corrected to annotated $\APP$}
$$
$$
\frac{a~=~a':\UU_n}{\T_n(a)~=~\T_n(a')}~~~~~~
\frac{a~=~a':\UU_n}{\sT_n^p(a)~=~\sT_n^p(a'):\UU_p}n<p
$$
$$
\frac{a:\UU_n~~~~~~b:\UU_n~(\T_n(a))}{\T_{n}~(\Pi^{n} a b) = \Pi~{\T_{n}(a)}~{\T_{n}(b)}}
~~~~~~~~\frac{a:\UU_n~~~~~~b:\UU_n~(\T_n(a))}{\sT_{n}^{p}~(\Pi^{n} a b) = \Pi^{p} \sT_{n}^{p}(a) \sT_{n}^{p}(b):\UU_p}n< p~
$$
$$
\frac{}{\T_{n}({\UU^{n}_m}) = \UU_{m}}m<n~~~~~~~~~~ \frac{}{\sT_{n}^{p}~(\UU^{n}_m) = \UU^p_m:\UU_p}m<n<p
$$
$$
\frac{a:\UU_m}{\T_n(\sT_{m}^{n}(a)) = \T_{m}(a)}n<m~~~~~~
\frac{a:\UU_m}{\sT_{n}^p(\sT_{m}^n(a)) = \sT_m^p(a):\UU_p}{m<n<p}
$$
\end{figure}

We define $\alpha:\Delta\rightarrow\Gamma$ by induction on $\alpha$.
We have $\pp:\Gamma.A\rightarrow\Gamma$ and $[a]:\Gamma\rightarrow \Gamma.A$ and
$\alpha^+:\Delta.A\alpha\rightarrow\Gamma.A$ if $\alpha:\Delta\rightarrow\Gamma$ and $\Gamma\vdash A~\type$.

\begin{lemma}\label{subst}
  The following rule is admissible: if $\Gamma\vdash J$ and $\alpha:\Delta\rightarrow\Gamma$
  then $\Delta\vdash J\alpha$.
\end{lemma}

Using the techniques developed in \cite{coquand:sophia,abel:wroclaw,coq18},
we can also show that $\Pi$ is one-to-one. An important consequence is closure under $\beta$-reduction.

\begin{lemma}
  If $\APP(A,B,\LAM(A',B',b),a):T$ then $b[a]:T$.
\end{lemma}

\begin{proof}
  We first get that $\LAM(A',B',b)$ has to be of type $\Pi~A~B$ and then that $\Pi~A~B~\conv~\Pi~A'~B'$.
  Using that $\Pi$ is one-to-one, we get $A=A'$ and $B=B'~(A)$. We then have $b:B~(A)$ and $T = B[a]$.
  We deduce $b[a]:B[a]$ by substitution and Lemma \ref{subst} and $b[a]:T$ by the type equality rule.
\end{proof}

\section{Rules for a type theory $T_2$ with Russell style universes}

The syntax is now for types and terms
$$
A,B,a,b~::=~v_i~|~c~a~|~\lambda~A~ b~|~\Pi~{A}~B~|~\UU_n
$$
Note that the application $c~a$ is not annotated by types, and that the abstraction $\lambda~A~ b$ only has one type argument, not two.

The typing and conversion rules are in Figures \ref{type-Russell} and \ref{conv-Russell}.

\begin{figure}
  \caption{Typing rules for Russell-style}\label{type-Russell}
$$
  \frac{A~\type}{v_0:A\pp~(A)}~~~~~~\frac{v_n:A~~~~B~\type}{v_{n+1}:A\pp~(B)}
$$
$$
\frac{A~\type~~~~~~B~\type~(A)}{\Pi~A~B~\type}
~~~~~~~\frac{b:B~(A)}{\lambda~A~ b:\Pi~A~B}~~~~~~~~
\frac{c:\Pi~A~B~~~~~~a:A}{c~a:B[a]}~~~~~~
$$     
$$
\frac{}{\UU_n~\type}~~~~~~
\frac{A:\UU_{n}}{A~\type}
~~~~~~
\frac{A:\UU_{n}~~~~~~B:\UU_n~(A)}
     {\Pi~A~B:\UU_{n}}
~~~~~~~
\frac{}{\UU_m:\UU_{n}}m<n
~~~~~~~
\frac{A:\UU_{n}}{A:\UU_{p}}n<p
$$
$$
\frac{}{\UU_n:\UU_{n+1}}~~~~~~~
\frac{A:\UU_{n}}{A:\UU_{n+1}}
\footnote{PD: old versions, should we keep them? Uniqueness of type?}
$$
\end{figure}

\begin{figure}
  \caption{Conversion rules for Russell-style}\label{conv-Russell}
$$
\frac{ a:A~~~~~~ A~ \conv~ B}{ a:B}~~~~~~~~~
\frac{ a ~\conv~a':A~~~~~~ A  ~\conv~ B}{ a ~\conv~a':B}
$$
$$
\frac{A~=~B~~~~~A~=~C}{B~=~C}~~~~~~~~~\frac{A~\type}{A~=~A}~~~~~~~~~
\frac{a~=~b:A~~~~~a~=~c:A}{b~=~c:A}~~~~~~~~~\frac{a:A}{a~=~a:A}
$$
$$
\frac{A~=~A'~~~~~~B~=~B'~(A)}{\Pi~A~B~=~\Pi~A'~B'}~~~~~~~~
\frac{c~=~c':\Pi~A~B~~~~~~a~=~a':A}{c~a~=~c'~a':B[a]}
$$
$$
%\frac{A~=~A'~~~~~~b~=~b':B~(A)}{\LAM~A~ b~=~\LAM~A'~ b':\Pi~A~B}~~~~~~~~
\frac{b:B~(A)~~~~~~~~ a:A}{ (\LAM~A~ b)~a  ~\conv~ b[a]:B[a]}
~~~~~~~
\frac{f\pp~v_0 = g\pp~v_0:B~(A)}{ f = g : \Pi~A~B}
$$
$$
\frac{A~=~A':\UU_n}{A~=~A'}~~~~~~
\frac{A~=~A':\UU_n}{A~=~A':\UU_p}n<p
$$
\end{figure}
\footnote{PD: added the two last rules}





\medskip

We have an operation which removes the annotation from $T_1$ to $T_2$.
$$
|\UU_n| = \UU_n~~~~~~~~|\Pi~A~B| = \Pi~|A|~|B|~~~~~~~~~|\T_n(a)| = |a|
$$
$$
|\APP(A,B,c,a)| = |c|~|a|~~~~~~|v_i| = v_i~~~~~~|\LAM(A,B,b)| = \lambda~|A|~|b|~~~~~|\sT_m^n(a)| = |a|
~~~~~~~|\UU^n_m| = \UU_m
$$
The following remark has a direct proof.

\begin{lemma}
  If $\Gamma\vdash J$ in $T_1$ then $|\Gamma|\vdash |J|$ in $T_2$.
\end{lemma}


\section{Equivalence between $T_1$ and $T_2$}

%% We can describe neutral and $\beta$-normal forms for the non annotated terms.
%% $$
%% k~::=~v_i~|~k~n~~~~~~~~~~~n~::=~\UU_l~|~\Pi~n~n~|~\lambda n~|~k
%% $$


\begin{lemma}
 % If $\Gamma\vdash t:A$ and $\Gamma\vdash u:A$ then $\Gamma\vdash t = u:A$.
  If $\Gamma\vdash t:A$ and $\Gamma\vdash u:B$ and $|t| = |u|$ then
  \begin{itemize}
    \item either $\Gamma\vdash A = B$
      and $\Gamma\vdash t = u:A$
    \item or $\Gamma\vdash A = \UU_k$ and $\Gamma\vdash B = \UU_l$ and $\Gamma\vdash \sT_k^m(t) = \sT_l^m(u):\UU_m$ with $m = \max(k,l)$.
  \end{itemize}
\end{lemma}

The proof relies on the normalization property. First, it follows that $|\Gamma | \vdash |t|:|A|$ and $|\Gamma | \vdash |u|:|B|$. Then by normalization in $T_2$, $| t |$ and $| u |$ have a common normal form $v$  and by subject reduction
$|\Gamma | \vdash v : |A|$ and $|\Gamma | \vdash v : |B|$. By inspecting the types of normal forms in $T_2$ we conclude that either $|\Gamma | \vdash |A| = |B|$ or $|\Gamma | \vdash |A| = \UU_k$ and $|\Gamma | \vdash |B| = \UU_l$ and $|\Gamma | \vdash v :\UU_m$ with $m = \max(k,l)$. In the first case,  $|A|$ and $|B|$ have a common normal form $V$ and there is $V_1$ such that $| V_1 |  = V$. Hence, $\Gamma \vdash V_1 = A = B$ (why?). In the second case, $A = \UU_k$ and $B = \UU_l$ and $\sT_k^m(t) = \sT_l^m(u):\UU_m$.

\begin{theorem}
  If $\Gamma\vdash J$ in the system $T_2$ then there exists a unique, up to conversion, judgement $\Gamma_1\vdash J_1$
  in the system $T_1$ such that $|\Gamma_1| = \Gamma$ and $|J_1| = J$.
\end{theorem}

\section{Addition of data types}

The proof extends to the corresponding systems with data types.

For the case of the list types $\List~A$ with constructors $\Nil$ and $\Cons$, we can define in the Russell system
$$
f~:~\List~\UU_0~\rightarrow~\List~\UU_1~~~~~~~~~~~
f~\Nil = \Nil~~~~~~f~(\Cons~x~xs) = \Cons~x~(f~xs)
$$
which can be lifted to in the Tarski version
$$
f~:~\List~\UU_0~\rightarrow~\List~\UU_1~~~~~~~~~~~
f~\Nil = \Nil~~~~~~f~(\Cons~x~xs) = \Cons~(\sT_0^1 x)~(f~xs)
$$
and is extensionally the identity function.
In the Russell system, in the context  $xs:\List~\UU_0$,
we do have $f~xs:\List~\UU_1$ but we do \EMP{not} have $xs:\List~\UU_1$

\medskip

     Similarly, the system $T_2$ is \EMP{not} closed under $\eta$-reduction. In the context
     $x:\UU_1\rightarrow\UU_1$, 
     we have $\lambda_{y:\UU_0}x~y:\UU_0\rightarrow\UU_2$ but we do \EMP{not} have
     $x:\UU_0\rightarrow\UU_2$.




\section{Initial models}

It is possible to present type theory with cumulative hierarchy of universe as a generalised algebraic theory.
The system $T_1$ can be used to present the initial/term model of this theory, following the method presented in \cite{Streicher:semtt}
(method which was formalised in the work \cite{brunerie:initiality}).
Our result shows that Russell's system $T_2$ can be used as well to present this initial/term model.



\bibliographystyle{plain}
\bibliography{refs}

\end{document}
